From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-31 paper.Rmd
WIP

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
#preamble: "\\usepackage[nomarkers]{endfloat}"

@@

Stochasticity is most often thought of as something which obscures patterns we are trying to infer.  This is a simple consequence of modeling: as discussed above, variables of interest to the processes in question are modeled explicitly (e.g. change in population density as it depends on the current density), everything else is just noise.  Most statistical inference in ecology is built solely to deal with this problem: pulling back the veil of noise to reveal processes of interest.  Unfortunately, given the high levels of variation and relatively limited data typical in ecological analyses, this can be a significant challenge, and even as statistical and computational advances provide more powerful and accessible inference tools, the task only gets harder.  For instance, few ecological concepts are more central than the notion of density dependence: that competition in some form or other should limit a species ability to grow without bound.  Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by a landmark study of @Knape2011.  Using time series from the Global Population Dynamics Database (GPDD) @Knape2011 show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of available time series.  Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesied deluge of streaming micro-sensors stapled to every organism provide enough signal to beat back the noise. Fortunately, this has not been the path taken.

@@

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) \\
\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \frac{\partial b(x) }{\partial x} - \frac{\partial d(x) }{\partial x}\right) \sigma^2 + b(x) + d(x) \label{bd}
\end{align}

where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates.  We can solve these equations at steady-state by setting both equations zero; e.g. for our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.  Note that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.

<!-- Master equation solution of birth-death model, from appendix-->

@@

Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes whose dynamics we are not explicitly modelling.  A population-level model

Given ecologists natural inclination towards modeling the biological rather than environmental aspects of a system, it will be no surprise that mechanistic descriptions of environmental stochasticity have remained far more fuzzy.  One step in this direction has been recognition that whatever the mechanisms, this variation is unlikely to result in purely white noise, as commonly postulated for mathematical convenience, but rather some degree of correlation, termed red noise, or possibly anti-correlation such as from seasonal or other oscillatory patterns, so called blue noise.

@@

@@

<!-- Figure 2: Quasi-cycles -->
```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.  Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations. Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
read_csv("appendixA/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() #+ 
#  labs(x="Time", 
#       y="Population density",
#       title="Noise induces quasi-cycles",
#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
```

@@

We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.

@@

This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.  The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:

$$U(x) = -\int_0^x f(y) \textrm{d} y $$ 
where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.  The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.  This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.  The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g. @Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).  This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.  (Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).  Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.  While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.  While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.

@@

As we saw in the equations from the system-size approximation, the variance and autocorrelation in question are technically properties of an ensemble, and should be computed over replicate simulations.  While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.  Instead,

>>>>

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"

#preamble: "\\usepackage[nomarkers]{endfloat}"

@@

Stochasticity is most often thought of as something which obscures patterns we are trying to infer.  This is a simple consequence of modeling: as discussed above, variables of interest to the processes in question are modeled explicitly (e.g. change in population density as it depends on the current density), everything else is just noise.  Most statistical inference in ecology is built solely to deal with this problem: pulling back the veil of noise to reveal processes of interest.  Unfortunately, given the high levels of variation and relatively limited data typical in ecological analyses, this can be a significant challenge, and even as statistical and computational advances provide more powerful and accessible inference tools, the task only gets harder.  For instance, few ecological concepts are more central than the notion of density dependence: that competition in some form or other should limit a species ability to grow without bound.  Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by @Knape2011.  Using time series from the Global Population Dynamics Database (GPDD) Knape & de Valpine show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of available time series.  Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesied deluge of streaming micro-sensors stapled to every organism provide enough signal to beat back the noise. Fortunately, this has not been the path taken.

@@

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + O(N^{-1})\\
\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x) \label{bd}
\end{align}

where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates, assuming the transitions (+/- 1 in this case) are much smaller than the system size ($N$ in the Levins model).  Note that the dynamics for the average population match the population-level equation^[up to fluctuations of order $\tfrac{1}{N}$. We can actually both observe and successfully approximate this higher-order correction from additional terms in the system size expansion to recover the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model -- see Appendix A for examples in simulations compared to predictions of the system size expansion].

\begin{align}
\frac{\ud n}{\ud t} = \underbrace{c n \left(1 - \frac{n}{N}\right)}_{\textrm{birth}} - \underbrace{e n}_{\textrm{death}}, \label{levins}
\end{align}

Note that like the equation for the mean, we have a dynamic equation for the time-evolution of the variance in Eq \eqref{bd}.  We can solve these equations at steady-state by setting both equations zero. For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.  We can also see that different ways of partitioning the same macroscopic equation into components of birth and death will create the same mean dynamics but with different levels of noise.  For instance, in a logistic model with $b = rn$ and $d = rn^2/N$, we instead find the variance at steady state is $\sigma^2 = K$; significantly larger than in the Levins model.

@@

Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes whose dynamics we are not explicitly modelling.  Once again, this can best be illustrated through an explicit formulation of the master equation.  For a generic population model that depends not only on population state $x$ but also on environmental factors $y$, it is possible to derive the steady-state variance (Appendix B) as:

$$\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d$$

where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the autocorrelation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.  For example, if extinction rate $e$ in our Levins' model is allowed to vary stochastically as Gaussian white noise with variance $\sigma_e^2$ and mean $e$, we find the expected variance introduced in the population to be:

$$\sigma_n^2 = \frac{e^2}{(c+e)^2} \sigma_e^2 + \frac{e}{c}N $$

Note that the contribution of environmental stochasticity is independent of system size and directly proprotional to the variance in the environment.  This derivation provides a useful illustration of the difference between demographic noise and environmental noise: the latter simply describes the variation introduced by some other dynamic variable that we are not modelling explicitly, and whose dynamics are independent of the state of our system, $x$.  The correlation time $\tau_c$, is really a shorthand for $\tfrac{1}{\partial_y g(y)}$.  Coulpling

@@

<!-- Figure 2: Quasi-cycles -->
```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.  Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations. Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
read_csv("appendixA/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

@@

We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.  This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.  The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:

$$U(x) = -\int_0^x f(y) \textrm{d} y $$

where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.  The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.  This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.  The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g. @Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).  This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.  (Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).  Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.  While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.  While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.

@@

@@

As we saw in the equations from the system-size approximation, the variance and autocorrelation in question are technically properties of an ensemble, and should be computed over replicate simulations.  While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.  If the environmental change is sufficiently slow relative to the measurements, one can assume that the system is close to its stationary state over a given window in time, computing these averages over a rolling time window as we illustrate here, rather than across an ensemble of replicates.
